---
title: "Getting started with deaviz"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with deaviz}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}


has_bench   <- requireNamespace("Benchmarking",  quietly = TRUE)
has_smacof  <- requireNamespace("smacof",        quietly = TRUE)
has_igraph  <- requireNamespace("igraph",        quietly = TRUE) &&
               requireNamespace("graphlayouts",  quietly = TRUE)
has_kohonen <- requireNamespace("kohonen",       quietly = TRUE)

knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 7,
  fig.height = 4.8,
  fig.align  = "center",
  dpi        = 96,
  out.width  = "85%"
)
```

`deaviz` turns the numbers behind a Data Envelopment Analysis (DEA), whether the inputs/outputs profiles or the computational outcome of various DEA models, into
plots: efficiency distributions, input/output relationships, an efficient
frontier representation, projection biplots, benchmarking networks, cross-efficiency maps, and
multi-period trajectories for panel data. This vignette walks through the typical workflow on
two bundled datasets.

The workflow has three steps that map onto the package's naming convention:

1. wrap your data in a **`dea_data()`** object (inputs, outputs, DMU labels);
2. **`compute_*()`** the quantities you need (efficiency, cross-efficiency,
   weights, self-organising maps);
3. **`plot_*()`** the result.

```{r load}
library(deaviz)
```

Most plots compute efficiency scores internally, which relies on the **`Benchmarking`**
package; a few embeddings/layouts use `smacof`, `igraph`/`graphlayouts`, or
`kohonen`. These are all *Suggests*: install them to reproduce every figure
below. (Where a suggested package is missing, the corresponding chunk is simply
skipped, so this vignette always builds.)

## The data object

`deaviz` ships `chinese_cities`, a classic cross-sectional DEA benchmark of 35
cities with three inputs and three outputs (Sueyoshi, 1992). `dea_data()`
records which columns
are inputs, which are outputs, and which identifies the DMU. Columns can be
given by name or by position.

So the `dea_data` object can be defined either by the input and output variable names as follows: 

```{r dea-data}
d <- dea_data(
  chinese_cities,
  inputs  = c("industrial_labour_force", "working_funds", "investments"),
  outputs = c("gross_industrial_output", "profit_and_tax", "retail_sales"),
  id      = "DMU"
)
d
```

or equivalently by the location of them: 

```{r dea-data-by-position, eval = FALSE}
d <- dea_data(
  chinese_cities,
  inputs  = 2:4,
  outputs = 5:7,
  id      = "DMU"
)

```


Everything downstream takes this `d` object. If your input/output columns are
prefixed `i_` / `o_`, `dea_data()` detects them automatically and you can skip
the `inputs`/`outputs` arguments.

## Efficiency scores

`compute_efficiency()` returns radial efficiency scores (plus peer weights and
multiplier weights). Returns-to-scale and orientation are arguments.

```{r efficiency, eval = has_bench}
eff <- compute_efficiency(x = d, rts = "vrs", orientation = "in")
head(round(eff$eff, 3))
```

## Distributions: the shape of efficiency

Start by looking at the spread. `plot_efficiency_distributions()` shows the
distribution of efficiency scores, while `plot_io_distributions()` shows the raw
input/output variables.

```{r dist, eval = has_bench}
plot_efficiency_distributions(d, rts = "vrs", title = "Chinese Cities Efficiency Scores", subtitle = "Variable Return To Scale")
```

```{r io-dist, eval = has_bench, fig.height = 5.2}
plot_io_distributions(d, type = "box", x_angle = 30)
```

Note the `x_angle = 30`: the input/output names are long, so tilting the x-axis
tick labels keeps them readable. Every plot whose x-axis carries variable or DMU
names accepts `x_angle`.

`plot_io_efficients()` counts how many DMUs are efficient versus inefficient.

```{r efficients, eval = has_bench, fig.height = 3.6}
plot_io_efficients(d, rts = "vrs", transparency = 1)
```

## Input/output relationships and the frontier

`plot_io_scatter()` lays out every input-against-output pair if no vector of inputs and/or outputs is assigned to `vars =` and if a vector provided, then the scatterplots will be limited to the pair-wise combination of the the variables. Regardless, the visual marks are coloured by
efficiency.

```{r scatter, eval = has_bench, fig.height = 6}

plot_io_scatter(d, vars = c("industrial_labour_force", "gross_industrial_output", "retail_sales") , color = "vrs")
```

It is possible to plot scatters against the efficiency scores of the DMUs as well: 

```{r scatter-efficiency, eval = has_bench, fig.height = 6}

plot_io_scatter(d, vars = c("industrial_labour_force", "gross_industrial_output", "retail_sales"), efficiency = "vrs" , color = "vrs")
```


The only frontier visualization plot is available via 
`plot_io_costa_frontier()` and it collapses all inputs and outputs onto a
single aggregated frontier (Bana e Costa et al., 2016).



```{r costa, eval = has_bench}
plot_io_costa_frontier(d)
```

## The textbook frontier

For teaching it is often clearest to take a single input against a single
output and draw the efficient frontier the way textbooks do.
`plot_io_frontier()` does exactly that: pick one input and one output, choose
the returns-to-scale, and it shades the production-possibility set and traces
the frontier, colouring each unit by its efficiency in that two-variable
sub-model.

```{r frontier, eval = has_bench}
plot_io_frontier(d, input = "working_funds", output = "retail_sales",
                 rts = "vrs")
```

Turn on `peers_network` to see which efficient peers an inefficient city is
benchmarked against; the arrows are weighted by the envelopment ($\lambda$)
weights.

```{r frontier-peers, eval = has_bench}
plot_io_frontier(d, input = "working_funds", output = "retail_sales",
                 rts = "vrs", peers_network = TRUE, labels = "Chengdu")
```

## Projection biplots

Two ways to project the multidimensional input/output space onto a readable
plane. `plot_io_pca_biplot()` uses PCA, drawing the input/output loading vectors;
`plot_io_mds()` uses metric (ratio) multidimensional scaling via the **smacof**
majorization algorithm (de Leeuw & Mair, 2009). The graphical use of such
projections for DEA follows Adler & Raveh (2008).

Let's have a look at the PCA biplot: 

```{r pca, eval = has_bench}
plot_io_pca_biplot(d, rts = "vrs")
```

The vectors are the dataset's inputs and outputs and they show the direction towards which the value of the corresponding input or output increases in the 2d space. 

In contrast, we can use an MDS algorithm to reduce the dimensionality of the dataset and represent the DMUs visually in a 2d plot. 

```{r mds, eval = has_bench && has_smacof}
plot_io_mds(d)
```

What to do with overcrowded plots? One solution offered by the `deaviz` package is to make the plot interactive so one can zoom into the plot and hover over the visual marks to get information about them. 

```{r mds-interactive, eval = FALSE}
plot_io_mds(d, interactive = TRUE)
```

## Benchmarking networks

For inefficient DMUs, DEA identifies the efficient peers they are benchmarked
against. `plot_io_lambda_network()` draws those peer relationships weighted by
the envelopment ($\lambda$) weights, laid out with Sammon mapping (Sammon,
1969) as in Porembski et al. (2005); `plot_io_peer_network()` lays out who is a
peer to whom, therefore the edges are directed from the inefficient units to their targets. 


```{r lambda, eval = has_bench}
plot_io_lambda_network(d, rts = "vrs")
```

```{r peer, eval = has_bench && has_igraph}
plot_io_peer_network(d, rts = "vrs")
```

It is sometimes important in the networks to focus on and highlight a specific DMU and `deaviz` package addresses that need via the `labels =` argument: 

```{r peer-label, eval = has_bench && has_igraph}
plot_io_peer_network(d,layout = "fr",labels = "Xian", rts = "vrs")
```

Pay attention that the default layout of the network is changed by providing a pre-defined value to the `layout =` argument. 

## Cross-efficiency

Cross-efficiency scores every DMU using every other DMU's optimal weights
(Doyle & Green, 1994). `compute_cross_efficiency()` builds the matrix, which `plot_cem_heatmap()`
displays. `plot_cem_unfolding()` unfolds the same matrix into a map of who rates
whom favourably (Ashkiani & Mar-Molinero, 2017), and `plot_cem_weights_heatmap()` shows the underlying weight
profiles.

```{r cem, eval = has_bench, fig.height = 6}
cem <- compute_cross_efficiency(d)

plot_cem_heatmap(cem, x_angle = 90)
```

```{r cem-unfold, eval = has_bench && has_smacof}
plot_cem_unfolding(cem)
```

```{r cem-weights, eval = has_bench, fig.height = 6}
plot_cem_weights_heatmap(d, x_angle = 30)
```

What if we want to highlight a specific DMU? Same as before, we can use the `labels=` argument: 

```{r cem-weights-focus, eval = has_bench, fig.height = 6}
plot_cem_weights_heatmap(d, x_angle = 30, labels = "Xian")
```

## Profile plots

`plot_io_radar()` and `plot_io_parcoo()` show each DMU's full input/output
profile, as a radar polygon or a parallel-coordinates line.

```{r radar, eval = has_bench, fig.width = 6.5, fig.height = 5.5}
plot_io_radar(d, efficiency = "vrs")
```

```{r parcoo, eval = has_bench, fig.height = 4.2}
plot_io_parcoo(d, efficiency = "vrs", x_angle = 30)
```

Both accept a `variables` argument to show a subset of the profile:
`"all"` (the default), `"inputs"`, `"outputs"`, or a vector of variable names
or positions.

```{r parcoo-variables, eval = has_bench, fig.height = 4.2}
plot_io_parcoo(d, variables = "inputs", efficiency = "vrs", x_angle = 30)
```

## Highlighting a single DMU: the focus view

This feature is so useful that it demands a dedicated section to explain it in further details. Pass one DMU name to `labels` and `deaviz` puts it centre stage: it is ringed and
labelled, while everything else fades back. For the network plots the focus keeps
the chosen DMU's sub-network; for the panel biplot it keeps that DMU's
trajectory.

```{r focus, eval = has_bench}
plot_io_pca_biplot(d, rts = "vrs", labels = "Beijing")
```

The amount of fade is tunable through the same `fade` argument: `TRUE` (default)
uses a sensible level, `FALSE` turns it off, and a number sets the alpha of the
faded marks directly (larger keeps them more visible).

```{r focus-level, eval = has_bench}
plot_io_parcoo(d, efficiency = "vrs", labels = "Beijing", fade = 0.4)
```

Other `labels` modes are `"all"` (label everyone), `"id"` (number each marker),
and `"max.overlaps"` (label as many as fit without collision).

```{r cem-unfold-id, eval = has_bench && has_smacof}
plot_cem_unfolding(cem, labels = "id")
```

## Self-organising maps

`compute_som()` trains a self-organising map (Kohonen, 2001) on the
input/output profiles, via the **kohonen** package (Wehrens & Kruisselbrink,
2018); `plot_io_som()` colours the map by mean efficiency per node.

```{r som, eval = has_bench && has_kohonen, fig.height = 5}
som <- compute_som(d)
plot_io_som(som)
```

## Multi-period data: trajectories

For panel data, `plot_panel_io_biplot()` places every DMU-period on a shared PCA
biplot and joins each DMU's points into a trajectory over time. The bundled
`taiwanese_banks` dataset is a balanced panel of 22 commercial banks over
2009-2011 (Kao & Liu, 2014). Inputs and outputs can be given by position.

```{r panel, eval = has_bench, fig.width = 7.5, fig.height = 5.5}
plot_panel_io_biplot(
  taiwanese_banks, id = "DMU", period = "Year",
  inputs = 3:5, outputs = 6:8, labels = "Cathay"
)
```

Here the focus view keeps only Cathay's three-year path lit while the other banks
recede, and the loading vectors are spread apart so their labels stay legible. 

It is noteworthy that the PCA is computed over the pooled data. 

For panel work it pays to build a `dea_panel_data()` object once -- it
validates the panel (unique DMU-period rows, balancedness) and remembers which
columns are the id, the period, and the inputs/outputs, so the `plot_panel_*`
functions can be called without repeating them:

```{r panel-data-object, eval = has_bench}
pd <- dea_panel_data(taiwanese_banks, inputs = 3:5, outputs = 6:8,
                     id = "DMU", period = "Year")
pd
```

`plot_panel_io_parcoo()` offers a simpler view of the same panel: the periods
form the parallel axes and each DMU is one line, tracing either a single
input/output variable or the per-period efficiency score (each period's
cross-section solved as its own DEA problem).

```{r panel-parcoo, eval = has_bench, fig.height = 4.6}
plot_panel_io_parcoo(pd, y = "vrs", labels = "Cathay")
```

With `color_by_trend = TRUE` each line is coloured by the DMU's overall change
(last period minus first), so the biggest risers and fallers stand out.

```{r panel-parcoo-trend, eval = has_bench, fig.height = 4.6}
plot_panel_io_parcoo(pd, y = "demand_deposits", color_by_trend = TRUE)
```

## Interactive plots

Many plots accept `interactive = TRUE`, returning a `plotly` widget with hover
tooltips instead of a static `ggplot`. This needs the `plotly` package and is
best viewed in an HTML context:

```{r interactive, eval = FALSE}
plot_io_pca_biplot(d, rts = "vrs", labels = "Beijing", interactive = TRUE)
```

## References

Adler, N., & Raveh, A. (2008). Presenting DEA graphically. *Omega*, 36(5),
715--729.

Ashkiani, S. (2019). *Four Essays on Data Visualization and Anomaly Detection of
Data Envelopment Analysis Problems* (PhD thesis). Universitat Autonoma de
Barcelona. <https://ddd.uab.cat/record/240333>

Ashkiani, S., & Mar-Molinero, C. (2017). Visualization of cross-efficiency
matrices using multidimensional unfolding. In *Recent Applications of Data
Envelopment Analysis*.

Bana e Costa, C. A., Soares de Mello, J. C. C. B., & Angulo Meza, L. (2016). A
new approach to the bi-dimensional representation of the DEA efficient frontier
with multiple inputs and outputs. *European Journal of Operational Research*,
255(1), 175--186. <https://doi.org/10.1016/j.ejor.2016.05.012>

Charnes, A., Cooper, W. W., & Rhodes, E. (1978). Measuring the efficiency of
decision making units. *European Journal of Operational Research*, 2(6),
429--444. <https://doi.org/10.1016/0377-2217(78)90138-8>

de Leeuw, J., & Mair, P. (2009). Multidimensional scaling using majorization:
SMACOF in R. *Journal of Statistical Software*, 31(3), 1--30.
<https://doi.org/10.18637/jss.v031.i03>

Doyle, J., & Green, R. (1994). Efficiency and cross-efficiency in DEA:
Derivations, meanings and uses. *Journal of the Operational Research Society*,
45(5), 567--578. <https://doi.org/10.1057/jors.1994.84>

Kao, C., & Liu, S.-T. (2014). Multi-period efficiency measurement in data
envelopment analysis: The case of Taiwanese commercial banks. *Omega*, 47,
90--98. <https://doi.org/10.1016/j.omega.2013.09.001>

Kohonen, T. (2001). *Self-Organizing Maps* (3rd ed.). Springer.

Porembski, M., Breitenstein, K., & Alpar, P. (2005). Visualizing efficiency and
reference relations in data envelopment analysis with an application to the
branches of a German bank. *Journal of Productivity Analysis*, 23(2), 203--221.
<https://doi.org/10.1007/s11123-005-1328-5>

Sammon, J. W. (1969). A nonlinear mapping for data structure analysis. *IEEE
Transactions on Computers*, C-18(5), 401--409.
<https://doi.org/10.1109/T-C.1969.222678>

Sueyoshi, T. (1992). Measuring the industrial performance of Chinese cities by
data envelopment analysis. *Socio-Economic Planning Sciences*, 26(2), 75--88.
<https://doi.org/10.1016/0038-0121(92)90015-W>

Wehrens, R., & Kruisselbrink, J. (2018). Flexible self-organizing maps in
kohonen 3.0. *Journal of Statistical Software*, 87(7), 1--18.
<https://doi.org/10.18637/jss.v087.i07>

## Where to next

Every function has its own help page with a full argument list and examples
(e.g. `?plot_panel_io_biplot`). To cite the package, see
`citation("deaviz")`.
